---
title: "RPMM Tutorial with Ordered Heat Map"
author: "David Chen, Christensen Lab"
maintainer: "David Chen, Christensen Lab"
email: "ydavidchen@dartmouth.edu"
date: "6/11/2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparation and Import Packages

We will use a built-in data set in the RPMM R package. No external data set is needed for this tutorial. 

```{r imports}
library(RPMM)
library(seriation)
library(pheatmap)

set.seed(1234) #reproducibility
data("IlluminaMethylation") #217 tissues x 100 CpGs; see documentation
Y_inv <- IllumBeta #copy
rownames(Y_inv) <- paste("Sample", rownames(Y_inv), sep="_")
```

For this demo, we will consider a simple scenario by spliting the data set only twice. To do so, we will set the `maxlevel` argument to the desired number. If not set, the algorithm defaults to $\infty$, i.e. no limit on the number of possible clusters.

```{r rpmm}
nSplit <- 2 #if not set, defaults to Inf
myRPMM <- blcTree(Y_inv, maxlevel=nSplit, verbose=1)
myRPMM
```

In the following plot, each row is the mean value of each methylation probe, and each column is an RPMM cluster of samples:

```{r}
plotImage.blcTree(myRPMM)
```

We will also examine the clustering dendrogram for RPMM:

```{r}
plotTree.blcTree(myRPMM)
title("RPMM Clustering Dendrogram")
```

## Extract RPMM cluster assignment

The RPMM function `blcTreeLeafClasses` extracts cluster assignments (hard cluster labels). The order of the samples are the same as the `row.names` of the matrix (which is `Y_inv` here) passed into the `blcTree` function. You may refer to the example from the RPMM package for the official implementation.

```{r labelExtracted}
myHardLabels <- blcTreeLeafClasses(myRPMM)
names(myHardLabels) <- rownames(Y_inv) 

myHardLabels <- as.data.frame(myHardLabels)
colnames(myHardLabels) <- "RPMM"
table(myHardLabels$RPMM)
```

## (Advanced) Extract the posterior probabilities for the cluster assigned

RPMM cluster assignment assumes a *softmax* framework: Each sample will be assigned probabilities ("soft labels") for all possible classes. A cluster assignment ("hard label") is determined based on the maximum probability.

Now extract posterior cluster probabilities as a matrix, and then manually apply *softmax*:

```{r posterior}
posteriorProbs <- blcTreeLeafMatrix(myRPMM)
rownames(posteriorProbs) <- rownames(Y_inv)

unsup_soft <- data.frame(
  RPMM.maxPosterior = apply(posteriorProbs, 1, FUN=max),
  RPMM.hard = colnames(posteriorProbs)[apply(posteriorProbs, 1, FUN=which.max)]
)
```

We can (and should) check to see if the hard probabilities we manually determined are the same as those returned by `blcTreeLeafClasses`:

```{r}
all( as.character(unsup_soft$RPMM.hard) == as.character(myHardLabels$RPMM)  )
```

## Heat Map Visualization

We are now ready to visualize the data. While many R packages could be used to draw heat maps, here we choose `pheatmap` as it is rather convenient to use. Other packages such as `heatmap.3` could also be used, though their syntax may vary. 

### Determine RPMM sample orders

We will use the `seriation` package to retrieve sample order. The detail of this procedure can be found in *pages 15-17 of the tutorial* authored by Hinoue et al. "TCGA COAD/READ DNA Methylation" (2012).

```{r seriation}
sampOrder <- c()
for( r in names(table(myHardLabels$RPMM)) ) {
  samps <- rownames(myHardLabels)[myHardLabels$RPMM == r]
  clu <- t(Y_inv[samps, ]) #Hinoue et al
  s_i <- seriate(clu, margin=2) #Hinoue et al
  so_i <- get_order(s_i) #Hinoue et al
  sampOrder <- c(sampOrder, samps[so_i]) #Hinoue et al
}
```

### Draw heat map with tracking bars

By default, `pheatmap` will implement hierarchical clustering with Euclidean distance and complete linkage (`hclust` in base; see documentation). Here, since we used RPMM to identify clusters and extracted sample orders from the RPMM output, we will:

* Turn off `pheatmap`'s clustering algorithm
* Re-order the data matrix based on sample order extracted from RPMMM
* Draw heat map based on RPMM sample order that we have extracted

Details regarding constructing the rows and tracking bars can be found in the `pheatmap` documentation and working examples, and won't be re-introduced in this tutorial.

```{r dataVis, fig.width=11.69, fig.height=8.27}
sample_annotation <- as.data.frame(myHardLabels) #or merge myHardLabels into existing annotation
mat <- t(Y_inv[sampOrder, ]) #invert to row=feature, column=sample format
pheatmap(
  mat = mat, #reorder based on sample order retrieved
  cluster_cols = FALSE, #turn off hierarchical clustering
  annotation_col = sample_annotation,
  color = colorRampPalette(c("yellow","black","blue"))(512),
  border_color = NA,
  fontsize_row = 5,
  fontsize_col = 4
)
```

## Other Notes

* You may proceed to testing cluster membership with available variables, such as by disease status (i.e. 1=cases vs. 0=controls) or the presence of a specific mutation among tumor samples (i.e. 1=mutated, 0=wild-type).

* In addition to DNA methylation, RPMM algorithm could be applied to a variety of molecular data types. The Gaussian RPMM implemented in the function `glcTree` may be useful.

## References

Hinoue T, Weisenberger DJ, Laird PW. TCGA COAD/READ DNA Methylation (2012) `https://pdfs.semanticscholar.org/f47a/28e8e2f7598b6e94de7ef3279957b6e55ec2.pdf`

Houseman EA, Christensen BC, Yeh RF, Marsit CJ, Karagas MR, Wrensch M, Nelson HH, Wiemels J, Zheng S, Wiencke JK, Kelsey KT. Model-based clustering of DNA methylation array data: a recursive-partitioning algorithm for high-dimensional data arising as a mixture of beta distributions. BMC Bioinformatics, 9(1): 365 (2008) `https://www.ncbi.nlm.nih.gov/pubmed/18782434/`

Houseman. RPMM: Recursively Partitioned Mixture Model. CRAN R package version 1.25 `https://cran.r-project.org/web/packages/RPMM/`

